Tracing the Hercules stream around the Galaxy

### 2010-05-20 tex/hercules.tex
groups-simulations: intro+references

baryon acoustic feature, coincidence problem

@@

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg,][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c = \vo(R/\Ro)^\beta \,,
\end{equation}
and for most of the simulations below we specify this model further to
a flat rotation curve ($\beta = 0$). The model for the bar is given by
the following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, following the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}

@@

As shown by \citet{dehnen99c,dehnen00a}, the model described above
that places the Sun at \Ro = 0.9\Rolr---where \Rolr\ is the radius at
which a circular orbit is in the outer Lindblad resonance with the
bar---at a current angle of 25$^{\circ}$ with the bar, with a bar
strength $\alpha = 0.01$, and integrating for four bar periods,
reproduces the Hercules feature in the local velocity distribution
while agreeing with other photometric and kinematical observations of
the bar. In what follows, unless indicated otherwise, we fix all of
the parameters of the model at these fiducial values.

@@

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the tangential velocity. This
\figurename\ is no longer symmetric with respect to shifts of
180$^{\circ}$ degree in azimuth, since the line-of-sight direction
breaks this symmetry. The gray curve is the line-of-sight distribution
one would observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric Galaxy
and the barred Galaxy. These predictions are also largely a function
of the particular form and parameters of the distribution function
that we chose the represent the stellar disk before bar formation.

@@

To further analyze the best directions to target to best see the
Hercules feature in the line-of-sight velocities, I have calculated
the line-of-sight velocity distribution on a 100 by 20 grid in
Galactocentric azimuth and radius. As a measure of the difference
between the barred Galaxy model and the steady-state, non-barred
model, I have calculated the Kullback-Leibler divergence between these
predicted distributions \citep[\eg,][]{mackay}
\begin{equation}
D_{\text{KL}}(P||Q) = \int \dd x\,p(x)\,\ln \frac{p(x)}{q(x)}\,,
\end{equation}
where, in this case, $x$ is the line-of-sight velocity, $p(x)$ is the
predicted distribution for the bar-origin model of the Hercules
stream, and $q(x)$ is the axisymmetric prediction. The
Kullback-Leibler divergence is only one among many possible choices to
quantify the difference between two probability distributions. It can
be interpreted as the expected weight of evidence for the prediction
by the barred Galaxy model over the fully axisymmetric prediction that
can be expected from each sampled star. As such, locations with the
maximal Kullback-Leibler divergence are those in which the Hercules
feature stands out most prominent in the line-of-sight.

@@

%\bibitem[{{Binney} \& {Tremaine}(2008)}]{binneytremaine}
%{Binney},~J. \& {Tremaine},~S., 2008, {Galactic Dynamics: Second Edition}
%  (Princeton University Press)

@@

\bibitem[Blitz \& Spergel(1991)]{blitz91a}
  Blitz,~L.~\& Spergel,~D.~N., 1991, \apj, 379, 631

@@

%\bibitem[Dehnen(1999a)]{dehnen99a}
%  Dehnen,~W., 1999a, \aj, 118, 1190

@@

\bibitem[H\"{a}fner \etal(2000)]{hafner00}
  H\"{a}fner,~R., \etal, 2000,
  \mnras, 314, 433

@@

\bibitem[Minchev \& Famaey(2009)]{minchev09a}
  Minchev,~I.~\& Famaey,~B., 2009,
  \apj L, submitted

@@

\bibitem[Sellwood(2010)]{sellwood10a}
  Sellwood,~J.~A., 2010,
  \mnras, submitted, 
  arXiv:1001.5197v1 [astro-ph.GA]

@@

\bibitem[Skuljan, Hearnshaw, \& Cottrell(1999)]{Skuljan99a}
  Skuljan,~J., Hearnshaw,~J.~B., \& Cottrell,~P.~L., 1999, \mnras, 308, 731

@@

\clearpage
\begin{figure}
\includegraphics[width=0.5\textwidth]{veldist.eps}
\caption{The velocity distribution of nearby ($\lesssim 100$ pc) stars
  from \citet{Bovy09a}. The Hercules moving group is visible as an
  overdensity at \vphi $\approx$ -0.2 $\vo$.}\label{fig:obs}
\end{figure}

>>>>

The velocity distribution of nearby stars contains various large
overdensities of co-moving stars that cannot be explained as the
vestiges of the clumpy nature of star formation
\citep[\eg,][]{Bovy10a}. In particular, the Hercules moving group, a
group of stars significantly offset from the bulk of the observed
velocity distribution \citep[\eg,][see
  \figurename~\ref{fig:obs}]{Dehnen98b,Bovy10a}, displays a wide range
of ages and metallicities
\citep{Blaauw70a,raboud98a,caloi99a,Bensby07a}. A compelling model
that has it originating through resonant interactions of stars in the
outer disk with the bar in the central region of the Galaxy has been
put forward \citep{dehnen00a,fux01a}, but any confirmation of this
picture beyond the locallly observed kinematics of stars remains
lacking. An investigation of the Hercules members' color--magnitude
diagram found no evidence for any significant metallicity anomaly with
respect to other local stars \citep{Bovy10a}, even though such a
difference should be expected as the Hercules stars originate from a
few kiloparsec toward the Galactic center where the average
metallicity is higher than in the Solar neighborhood. The absence of
the predicted metallicity anomaly could potentially be explained
through, \eg, radial mixing \citep{sellwood02a}, but it is clear that
chemodynamical modeling of the Galactic disk is currently not up to
the task of separating out these effects.

In this Letter I propose that a robust and unambiguous test of the
bar-resonance model for the Hercules stream is the distinct pattern it
predicts as we trace the velocity distribution of stars near the Solar
radius around the Galaxy. \Gaia\ is in a good position to have the
final word on this, but I also show that the Hercules feature can be
detected in the line-of-sight velocity distribution in selected
directions on the sky. Clear signatures should exist in regions a few
kiloparsec from the Sun, allowing for a clean verification of the
bar-resonance model for the Hercules stream.

@@

Since the Hercules stream is only apparent in the planar motions in
the disk \citep[\eg,][]{Bovy09a}, we only consider a two-dimensional
model for the Galaxy. We use a simple power-law rotation curve to
model the axisymmetric Galactic potential
\begin{equation}
v_c = \vo(R/\Ro)^\beta \,,
\end{equation}
and for most of the simulations below we specify this model further to
a flat rotation curve ($\beta = 0$). The model for the bar is given by
the following potential
\begin{equation}
\Phi_b(R,\phi) = \Ab(t) \cos [2(\phi - \Omegab t)] \times
\left\{ \begin{array}{ll} -(\Rb/R)^3\,, & \mathrm{for}\ R \geq
  \Rb\,,\\ (R/\Rb)^3-2\,, & \mathrm{for}\ R \leq \Rb\,. \end{array}
\right.
\end{equation}
Here \Omegab\ is the pattern speed of the bar and \Rb\ is the bar
radius, which is fixed to be 80\,percent of the bar's corotation
radius. The bar is grown smoothly during a time $t_1$, always set to
half of the total integration time in what follows, following the
prescription
\begin{equation}
\Ab(t) = \alpha \, \frac{\vo^2}{3}\,\left(\frac{\Ro}{\Rb}\right)^3
\times \left\{ \begin{array}{lll}
\frac{3}{16}\xi^5 - \frac{5}{8} \xi^3 + \frac{15}{16} \xi + \frac{1}{2}\,, & \xi \equiv 2\frac{t}{t_1}-1\,, & t \leq t_1\,,\\
1\,, & & t \geq t_1\,.\end{array} \right. 
\end{equation}
A comparison of the local velocity distribution obtained by using this
simple bar potential with that resulting from the more realistic bar
potential of \citet{gardner10a} shows that this simple model is
sufficient to model the bar's influence on stellar orbits in the outer
disk.

@@

As shown by \citet{dehnen99c,dehnen00a}, the model described above
that places the Sun at \Ro = 0.9\Rolr---where \Rolr\ is the radius at
which a circular orbit is in the outer Lindblad resonance with the
bar---at a current angle of 25$^{\circ}$ with the bar, with a bar
strength $\alpha = 0.01$, and integrating for four bar periods,
reproduces the Hercules feature in the local velocity distribution
while agreeing with other photometric and kinematical observations of
the bar \citep[\eg,][]{binney97a,bissantz02a,cole02a,Shen10a}. In what
follows, unless indicated otherwise, we fix all of the parameters of
the model at these fiducial values.

@@

In \figurename~\ref{fig:rphi1d} I show the predicted line-of-sight
velocity distributions in various locations near the Solar circle all
around the Galaxy, obtained by integrating the full two-dimensional
velocity distribution over the tangential velocity. This
\figurename\ is no longer symmetric with respect to shifts of
180$^{\circ}$ degree in azimuth, since the line-of-sight direction
breaks this symmetry. The gray curve is the line-of-sight distribution
one would observe if the targeted stars were in a steady state in an
axisymmetric potential with a Dehnen distribution function
(\eqnname~[\ref{eq:fdehnen}]). Both line-of-sight velocity
distributions are normalized to have equal mass since I assume that
our knowledge of the Galaxy and the tracer population will not be
great enough that we can test predicted differences in the absolute
number of stars at a given location between the axisymmetric and the
bar-resonance predictions. These predictions are also largely a
function of the particular form and parameters of the distribution
function that we chose the represent the stellar disk before bar
formation.

@@

To further analyze the best directions to target to best see the
Hercules feature in the line-of-sight velocities, I have calculated
the line-of-sight velocity distribution on a 100 by 20 grid in
Galactocentric azimuth and radius. As a measure of the difference
between the bar-resonance model and the steady-state, axisymmetric
model, I have calculated the Kullback-Leibler divergence between these
predicted distributions \citep[\eg,][]{mackay}
\begin{equation}
D_{\text{KL}}(P||Q) = \int \dd x\,p(x)\,\ln \frac{p(x)}{q(x)}\,,
\end{equation}
where, in this case, $x$ is the line-of-sight velocity, $p(x)$ is the
predicted distribution for the bar-origin model of the Hercules
stream, and $q(x)$ is the axisymmetric prediction. The
Kullback-Leibler divergence is only one among many possible choices to
quantify the difference between two probability distributions. It can
be interpreted as the expected weight of evidence for the prediction
by the bar-resonance model over the fully axisymmetric prediction that
can be expected from each sampled star. As such, locations with the
maximal Kullback-Leibler divergence are those in which the Hercules
feature stands out most prominent in the line-of-sight.

@@

@@

@@

@@

@@

@@

@@

@@

\clearpage
\begin{figure}
\includegraphics[width=0.5\textwidth]{veldist.eps}
\caption{The velocity distribution of nearby ($\lesssim 100$ pc) stars
  from \citet{Bovy09a}. \vR\ and \vphi\ are the velocities toward the
  Galactic center and in the direction of Galactic rotation,
  respectively. The Hercules moving group is visible as an overdensity
  at \vphi $\approx$ -0.2 $\vo$.}\label{fig:obs}
\end{figure}
